# Coding "Best" Practices

This training module was developed by Dr. Kyle R. Roell and Dr. Julia E. Rager

Spring 2023


<!--"Coding best practices - what do we want at the top of every script file with a date, what you are doing, goal of analysis, annotation, etc
Point to coding style best practices resources, and everyone has their own coding style but here are recommended best practices" -->

<!--"These haven't been made, but Julia has opinions, and PREHEAT has some coding style best practices - Kyle, then Alexis to review" -->


## Introduction to Training Module

In this training module, we will be going over coding "best" practices. The reason we have put "best" in quotes is because although these are we consider best or better practices, everyone has different coding styles, annotation styles, etc. Here, we hope to just give you a sense of what we do when coding, why we do it, and why we think it is important. We will also be pointing out other guides to style, annotations, and best practices that we suggest implementing into your own coding and that we try to follow, as well.

<b>Some of the questions we hope to answer in this module are:

+ What should I name my script?
+ What should I put at the top of every script and why is it important?
+ How should I annotate my code?
+ Why are annotations important?
+ How do I implement these coding practices into my own code?
+ Where can I find other resources to help with coding best practices?
</b>

In the following sections, we will be attempting to answer these questions. Keep in mind that the advice and suggestions in this section are just that: advice and suggestions. So please take them into consideration and integrate them into your own coding style as appropriate.

## Naming the Script

The first thing we need to talk about, which is sometimes overlooked in the discussion of coding practices, is script naming conventions and high level descriptive headers within a script. It is important to remember to name your script something concise, but descriptive. You want to be able to easily recognize what the script is for and does without a cumbersome, lengthy title. Some tips for script naming conventions:

+ Be concise, but descriptive
+ Consider using dates when appropriate 
+ It is common to use underscores, but generally avoid other special characters
+ Prefer using full words if possible, avoid non-standard acronyms

Keep in mind, each script should have a clear purpose within a given project. And, it is sometimes necessary, and often common, to have multiple scripts within one project that all pertain to different parts of the analysis. 

## Script Header

Once your script is created and named, it is generally a good idea to have a script header. The script header is responsible for defining:

+ Name - Longer or more readble name than script name.
+ Author(s) - Who wrote the script?
+ Date - When was the script developed?
+ Description - Give a more detailed description of the purpose of the script, also a chance to provide any notes or special considerations for this particular script.

In, R, it is common to include multiple "#", the comment operator, or a "#" followed by another special character, to start and end a block of coding annotation. This is common in starting the script and header of the file. An example of this, and a common header, may look something like this:

```{r}

########################################################################
########################################################################
### Script Longer Title
### 
### Description of what this script does!
### Also can include special notes or anything else here.
###
### Created by: Kyle Roell and Julia Rager
### Last updated: 01 May 2023
########################################################################
########################################################################


```


It is not necessary to follow this example exactly, but this is just to give an example of what a common header file may look like. We will now take a look at how annotations within the script itself can make a huge difference in understanding the code within.

## Annotations

Before we actually talk about coding style, it is important to address annotating yoru code. So, what are annotations and why are they important? Annotations are basically just notes embeded within your code, as comments that will not be run. The beauty of annotating your code is that you not only others, but future you, will be able to read through and better understand what a particular piece of code does. I cannot tell you how many times I have gone back to my code and thought, I wish I had annotated this better. Because of this, I suggest annotating your code while you go and in a very descriptive fashion. While not every single line needs an annotation, or a very detailed one, it is a good idea to provide comments and annotation as much as you can. 

#### General Annotation Style

In general, annotations will be commented out, descriptive short sentences that provide a short description of what you are doing or why you are doing something specific. This can be helpful when you are defining a covariate a specific way, performing a specific analytical technique, or just generally explaining why you are doing what you're doing.

```{r, eval=F}

# Performing logistic regression to assess assocation between xyz and abc
# Regression confounders: V1, V2, V3 ...

xyz.regression.output = glm(xyz ~ abc + V1 + V2 + V3, family=binomial(), data=example.data)

```

#### Mid-Script Headings

Another common approach to annotations is to use mid-script type headings to separate out the script into various sections. For example, you might want to create distinct sections for "Loading Packages, Data, and Setup", "Covariate Definition", "Correlation Analysis", "Regression Analysis", etc. This can help you, and others looking at your script, to ater find your way through the script more easily. It also can be more visually please to see the script split up into multiple sections as opposed to one giant chunk of code interspersed with comments.

```{r, eval=F}

###########################################################################
###########################################################################
###
### Regression Analyses
###
### You can even add some descriptions or notes here about this section!
###
###########################################################################


# Performing logistic regression to assess assocation between xyz and abc
# Regression confounders: V1, V2, V3 ...

xyz.regression.output = glm(xyz ~ abc + V1 + V2 + V3, family=binomial(), data=example.data)

```


<b>General tips for annotations:</b>

+ Make comments that are useful and meaningful, you don't need to comment every single line
+ In general, you probably won't overannotate your script, so more is generally better
+ That being said, don't write super long paragraphs every few lines
+ Split up your script into various sections using mid-script headings, as show above, when appropriate


#### Quick, Short Comments and Annotations

While it is important to provide descriptive annotations, not every one needs to be a sentence or longer. As stated previously, it is not necessary to comment every single line, but when you do want to add some brief annotations just to simply describe what is happening, it can be useful to include a brief comment.

```{r, eval=F} 

# Loading necessary packages

library(ggplot2)  # Plotting package

```

In the example above, we can see that this may be an appropriate, short annotation before starting to load the packages necessary for the script. Short, one line annotations can also be placed after lines to clarify that specific line. Short, quick annotations like this can be useful within the larger mid-script headings to sort of split up these larger sections of code. 


## Coding Style

Coding style if often a contentious topic! There are MANY styles of coding, and no two coders have the same exact style, even if following the same reference. Here, we will provide some guides to coding style and go over some of the basic, general tips for making your code readable and efficient.

```{r, eval=F}

# Example of using spacing for alignment of variable assignment

Longer_variable_name_x = 1
Short_name_y           = 2

# Note guides will suggest you use "<-" as the assignment operator
# There are many reasons for this, in general, for most situations, "<-" and "=" will do the same thing



# Spacing around certain , ex: "if ()", and operators
# Open curly brace should not be on its own line, but closing one should

# Example of poor style

if(Longer_variable_name_x <Short_name_y) {
  Short_name_y = 0}

# Example of better style

if (Longer_variable_name_x < Short_name_y) {
  Short_name_y = 0
}


```

<b>General tips for coding style:</b>

+ Variable names
  + Make them intuitive, short, but descriptive
  + Use the same convention throught (ex: separating words with "." or "_")
+ Separate long longs onto two or more (typically for loops or functions)
+ Use "<-" for assignment operator
+ Using spacing appropriately for readability
  + Alignment of lines
  + After certain keywords when appropriate
+ Be consistent throughout 
  + Ex: if you use Tidyverse conventions, continue to use it throughout your script
+ Try to make your code as readable as possible


<b>Style guides: </b>

+ [Google's Style Guide](https://web.stanford.edu/class/cs109l/unrestricted/resources/google-style.html)
+ [Tidyverse Style Guide](https://style.tidyverse.org)
+ [R Blogger's Best Coding Practices](https://bookdown.org/content/d1e53ac9-28ce-472f-bc2c-f499f18264a3/)




## Script Organziation

Lastly, it is important to note that organzing your script efficient can help with readabilty as well. In general, as stated before, the beginning of your script should contain some sort of header lines to describe the script. Here, we will go over the basic ordering we suggest for most scripts.
<b>

1. Header section
2. Loading libraries and data
3. Function defintions (if any user defined functions exist)
4. Data and variables manipuliation
5. Analyses

</b>

While following this exact organziation isn't absolutley, it can help people follow your code.


## Concluding Remarks

In this module we set out to help with coding style and best practices. And, while, there are many different coding styles, hopefully this brief guide helped demonstrate some of the basic concepts and practices. Please reference some of the style guides in the previous section, "Coding Style", and create your own style from best practices that works for you. Try to make your code understandable, readable, and consider how a future reader, including yourself, might go through your code and what they will need to know to be able to best follow it. 






