# Introduction to Statistical Tests - Test Your Knowledge Key
```{r}
library(tidyverse) 
library(ggpubr) 
```

```{r}
# Loading and viewing data
full.data <- read.csv("Module3_4/Module3_4_FullDemoChemData.csv")

head(full.data)
```


Let's first assess normality of birth weight (`BW`) using a histogram, a Q-Q plot, and the Shapiro-Wilk test.
```{r}
gghistogram(data = full.data, x = "BW", bins = 20)
ggqqplot(full.data$BW, ylab = "BW")
```


```{r}
shapiro.test(full.data$BW)
```

From the plots and the Shapiro-Wilk test, birth weight is considered to be normally distributed. Since we're interested in determining if there's statistical significance between a continuous variable (birth weight) and a categorical variable (maternal education), we'll use an ANOVA test.

<br>

Let's start with visualizing the distribution of birth weight across the maternal education categories.
```{r}
ggboxplot(data = full.data, x = "MEdu", y = "BW")
```

There doesn't seem to much a difference, but we'll go ahead and run the ANOVA test.
```{r}
anova(aov(data = full.data, BW ~ MEdu))
```

Given that the p value ~0.14, there are no significant differences in birth weight based on a mother's education level.
