# Database Integration - Test Your Knowledge Solution

```{r results=FALSE, message=FALSE}
library(sf)
library(tidyverse)
library(ggthemes)
```


```{r}
# Reading in files
load("Module7_4_InputData1.RData")

# Viewing geography, pollutant, AQS (confounder) data 
head(counties_shapefile)
head(epa_ap_county)
head(county_health)
```

## TYK Question 1

Is there an association between long-term, ambient NO~2~ concentrations and age-adjusted mortality at the county level (i.e., the crude model results)?

First transforming the pollutant data from a long to wide format. 
```{r}
epa_ap_county <- epa_ap_county %>% 
  select(-Units.of.Measure) %>% 
  pivot_wider(names_from = Parameter.Name, values_from = County_Avg)

head(epa_ap_county)
```

Note that the `NO2` contains a lot of missing data. However, we can move forward by merging the geography and pollutant data by the `State_County_Code` and the `County.Code`.
```{r}
# Merging mortality and air pollution data
model_data <- merge(epa_ap_county, cdc_mortality, by.x = "State_County_Code", by.y = "County.Code")

head(model_data)
```

Now running the linear regression model between NO~2~ concentrations and age-adjusted mortality.
```{r}
# running the linear regression model
m <- lm(Age.Adjusted.Rate ~ NO2,
        data = model_data, weights = 1/model_data$Age.Adjusted.Rate.Standard.Error) 

# viewing the model results through the summary function
summary(m)   
```
**Answer**: Based on these model results, there may indeed be an association between NO~2~ concentrations and age-adjusted mortality (p value = 0.09)

<br>

## TYK Question 2 

After adjusting for covariates, is there an association between long-term, ambient NO~2~ concentrations and age-adjusted mortality at the county level (i.e., the adjusted model results)?
```{r}
# First we merge the covariate data in with the AQS data
model_data <- merge(model_data, county_health, by.x = "State_County_Code", by.y = "County.5.digit.FIPS.Code", all.x = TRUE)

head(model_data)
```

```{r}
# Now we add some relevant confounders to the linear regression model
m <- lm(Age.Adjusted.Rate ~ NO2 + Adult.smoking + Adult.obesity + Food.environment.index + Physical.inactivity +
          High.school.graduation + Some.college + Unemployment + Violent.crime + Percent.Rural + Percent.Females +
          Percent.Asian + Percent.Non.Hispanic.African.American + Percent.American.Indian.and.Alaskan.Native + Percent.NonHispanic.white,
        data = model_data, weights = 1/model_data$Age.Adjusted.Rate.Standard.Error)

# And finally we check to see if the statistical association persists
summary(m)
```

**Answer**: After adjusting for confounders, there is now a statistically significant relationship between NO~2 ambient exposure and age-adjusted mortality.