# Normality Testing - Test Your Knowledge Solution

Prepare workspace
```{r message = FALSE}
# Load packages
library(openxlsx) # for importing data
library(tidyverse) # for manipulating and plotting data
library(ggpubr) # for making Q-Q plots with ggplot 

# Set theme for graphing
theme_set(theme_bw())
```

Import data
```{r message = FALSE}
# Import data
tyk_data <- read.xlsx("Module3_3_TYKInput.xlsx")
```

## TYK Question 1
Are any variables normally distributed in the raw data?

### Assess Raw Data

#### Histograms
```{r}
# Prepare data frame
tyk_data_long <- tyk_data %>%
  pivot_longer(!S_ID, names_to = "variable", values_to = "value")

# Make figure panel of histograms
ggplot(tyk_data_long, aes(value)) +
  geom_histogram(fill = "gray40", color = "black", binwidth = function(x) {(max(x) - min(x))/25}) +
  facet_wrap(~ variable, scales = "free") +
  labs(y = "# of Observations", x = "Value")
```

#### Q-Q Plots
```{r fig.align = 'center'}
ggqqplot(tyk_data_long, x = "value", facet.by = "variable", ggtheme = theme_bw(), scales = "free")
```

#### Shapiro-Wilk
```{r}
# Apply Shapiro Wilk test
shapiro_res <-  apply(tyk_data %>% select(-S_ID), 2, shapiro.test)

# Create results data frame
shapiro_res <- do.call(rbind.data.frame, shapiro_res)

# Clean data frame
shapiro_res <- shapiro_res %>% 
  
  ## Add normality conclusion
  mutate(normal = ifelse(p.value < 0.05, F, T)) %>%
  
  ## Remove columns that do not contain informative data
  select(c(p.value, normal)) 

# View cleaned up data frame
shapiro_res
```
**Answer**: There are no variables that are normally distributed. 

<br>

## TYK Question 2
Does psuedo log~2~ transforming the values makes the distributions overall more or less normally distributed?

### Assess Psuedo log~2~ Transformed Data
```{r}
# Apply psuedo log2 transformation to chemical data
tyk_data_pslog2 <- tyk_data %>%
  mutate(across(DEET:Azoxystrobin, ~ log2(.x + 1)))
```

#### Histograms 
```{r}
# Pivot data longer
tyk_data_pslog2_long <- tyk_data_pslog2 %>%
  pivot_longer(!S_ID, names_to = "variable", values_to = "value")

# Make figure panel of histograms
ggplot(tyk_data_pslog2_long, aes(value)) +
  geom_histogram(fill = "gray40", color = "black", binwidth = function(x) {(max(x) - min(x))/25}) +
  facet_wrap(~ variable, scales = "free") +
  labs(y = "# of Observations", x = "Value")
```

#### Q-Q Plots
```{r fig.align = 'center'}
ggqqplot(tyk_data_pslog2_long, x = "value", facet.by = "variable", ggtheme = theme_bw(), scales = "free")
```

#### Shapiro-Wilk
```{r}
# Apply Shapiro Wilk test
shapiro_res_pslog2 <-  apply(tyk_data_pslog2 %>% select(-S_ID), 2, shapiro.test)

# Create results data frame
shapiro_res_pslog2 <- do.call(rbind.data.frame, shapiro_res_pslog2)

# Clean data frame
shapiro_res_pslog2 <- shapiro_res_pslog2 %>% 
  
  ## Add normality conclusion
  mutate(normal = ifelse(p.value < 0.05, F, T)) %>%
  
  ## Remove columns that do not contain informative data
  select(c(p.value, normal)) 

# View cleaned up data frame
shapiro_res_pslog2
```
**Answer**: Psuedo log~2~ transforming makes the distributions more normally distributed.

<br>

## TYK Question 3
What are the average Kruskal-Wallis p-values for the raw and psudeo log~2~ transformed data?

### Calculate Overall Kruskal-Wallis P-Values
```{r}
# Raw data
mean(shapiro_res$p.value)

# Log2 data
mean(shapiro_res_pslog2$p.value)
```
**Answer**: The p-value for the raw data is 5.05 x 10^-18 and the p-value for the psuedo log~2~transformed data is 0.0043.
