# Normality Testing - Test Your Knowledge Key

## Prepare workspace

```{r message = FALSE}
# Load packages
library(openxlsx) # for importing data
library(tidyverse) # for manipulating and plotting data
library(ggpubr) # for making Q-Q plots with ggplot 

# Set theme for graphing
theme_set(theme_bw())
```

## Import data

```{r message = FALSE}
# Import data
tyk_data <- read.xlsx("Module3/TYK_Input_Data.xlsx")
```

## Assess raw data

### Histograms
```{r}
# Prepare data frame
tyk_data_long <- tyk_data %>%
  pivot_longer(!S_ID, names_to = "variable", values_to = "value")

# Make figure panel of histograms
ggplot(tyk_data_long, aes(value)) +
  geom_histogram(fill = "gray40", color = "black", binwidth = function(x) {(max(x) - min(x))/25}) +
  facet_wrap(~ variable, scales = "free") +
  labs(y = "# of Observations", x = "Value")
```

### Q-Q Plots
```{r fig.align = 'center'}
ggqqplot(tyk_data_long, x = "value", facet.by = "variable", ggtheme = theme_bw(), scales = "free")
```

### Shapiro-Wilk
```{r}
# Apply Shapiro Wilk test
shapiro_res <-  apply(tyk_data %>% select(-S_ID), 2, shapiro.test)

# Create results data frame
shapiro_res <- do.call(rbind.data.frame, shapiro_res)

# Clean data frame
shapiro_res <- shapiro_res %>% 
  
  ## Add normality conclusion
  mutate(normal = ifelse(p.value < 0.05, F, T)) %>%
  
  ## Remove columns that do not contain informative data
  select(c(p.value, normal)) 

# View cleaned up data frame
shapiro_res
```
## Assess log2-transformed data
```{r}
# Apply log2 transformation to chemical data
tyk_data_log2 <- tyk_data %>%
  mutate(across(DEET:Azoxystrobin, ~ log2(.x + 1)))
```

### Histograms 

```{r}
# Pivot data longer
tyk_data_log2_long <- tyk_data_log2 %>%
  pivot_longer(!S_ID, names_to = "variable", values_to = "value")

# Make figure panel of histograms
ggplot(tyk_data_log2_long, aes(value)) +
  geom_histogram(fill = "gray40", color = "black", binwidth = function(x) {(max(x) - min(x))/25}) +
  facet_wrap(~ variable, scales = "free") +
  labs(y = "# of Observations", x = "Value")
```

### Q-Q Plots
```{r fig.align = 'center'}
ggqqplot(tyk_data_log2_long, x = "value", facet.by = "variable", ggtheme = theme_bw(), scales = "free")
```

### Shapiro-Wilk
```{r}
# Apply Shapiro Wilk test
shapiro_res_log2 <-  apply(tyk_data_log2 %>% select(-S_ID), 2, shapiro.test)

# Create results data frame
shapiro_res_log2 <- do.call(rbind.data.frame, shapiro_res_log2)

# Clean data frame
shapiro_res_log2 <- shapiro_res_log2 %>% 
  
  ## Add normality conclusion
  mutate(normal = ifelse(p.value < 0.05, F, T)) %>%
  
  ## Remove columns that do not contain informative data
  select(c(p.value, normal)) 

# View cleaned up data frame
shapiro_res_log2
```

## Calculate overall Kruskal-Wallis p-values
```{r}
# Raw data
mean(shapiro_res$p.value)

# Log2 data
mean(shapiro_res_log2$p.value)
```

## TYK answers

1. Are any variables normally distributed in the raw data? **No.**

2. Does log2-transforming the values makes the distributions overall more or less normally distributed? **More normally distributed.**

3. What are the average Kruskal-Wallis p-values for the raw and log2-transformed data? **The p-value for the raw data is 5.05 x 10^-18 and the p-value for the log2-transformed data is 0.0043.**

