# Improving Data Visualizations - Test Your Knowledge Key


```{r}
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
```

```{r}
# Load the data
smoke_data <- read.csv("Module3_2/Module3_2_DataforVisualizations.csv")

head(smoke_data)
```

Like we did in the module, we'll start by changing the dataframe into a long format and creating a new column for scaled chemical concentrations.

```{r}
scaled_longer_smoke_data = pivot_longer(smoke_data, cols = 4:13, names_to = "Biomass_Burn_Condition", 
                                 values_to = "Chemical_Concentration") %>%
    # scaling within each chemical
    group_by(Chemical) %>%
    mutate(Scaled_Chemical_Concentration = scale(Chemical_Concentration)) %>%
    ungroup()

head(scaled_longer_smoke_data)
```

Although the next code block appears to be the same as the module it is not! This time, the underscore between the biomass and the burn condition is retained. For example, in the dataframe we have the class `Red_Oak_Smoldering`. In the code below, that will be changed to `Red Oak_Smoldering`. 

This allows us to the underscores as a separator to split the `Biomass_Burn_Condition` column into `Biomass` and `Burn_Condition` using the `separate` function later on.

```{r}
scaled_longer_smoke_data = scaled_longer_smoke_data %>%
    mutate(Biomass_Burn_Condition = ifelse(Biomass_Burn_Condition == "Red_Oak_Smoldering", "Red Oak_Smoldering",
                ifelse(Biomass_Burn_Condition == "Red_Oak_Flaming", "Red Oak_Flaming",
                    ifelse(Biomass_Burn_Condition == "Pine_Needles_Smoldering", "Pine Needles_Smoldering",
                        ifelse(Biomass_Burn_Condition == "Pine_Needles_Flaming", "Pine Needles_Flaming",
                            Biomass_Burn_Condition)))))


head(scaled_longer_smoke_data)
```

Splitting `Biomass_Burn_Condition` into `Biomass` and `Burn_Condition`.

```{r}
scaled_longer_smoke_data = scaled_longer_smoke_data %>%
    separate(Biomass_Burn_Condition, c("Biomass", "Burn_Condition"), sep = "_")

head(scaled_longer_smoke_data)
```

Putting `Burn_Condition` into a factor to reorder the classes for plotting purposes.

```{r}
scaled_longer_smoke_data$Burn_Condition = factor(scaled_longer_smoke_data$Burn_Condition, 
                                                 levels = c("Smoldering", "Flaming"))
```

## Heat Map

```{r}
options(repr.plot.width = 16, repr.plot.height = 4)

# selecting palette
rcolorbrewer_colors = brewer.pal(n = 3, name = 'Dark2')

Module3_2_TYK_figure = ggplot(data = scaled_longer_smoke_data, aes(x = Chemical.Category, y = Biomass,
                                           fill = Scaled_Chemical_Concentration)) + 
  geom_tile(color = 'white') + # adds white space between the tile

  facet_wrap(~Burn_Condition, scales = 'free_y') + 

  theme_light() + # changing the theme
  theme(strip.text = element_text(size = 18, face = "bold"), # changes size of facet axes 
      axis.text.x = element_text(angle = 20, hjust = 0.85), # rotating x labels/ moving slightly to the left
      axis.text = element_text(size = 12), # changing size of axis labels
      axis.title = element_text(face = "bold", size = rel(1.7)), # changes axis titles
      legend.title = element_text(face = 'bold', size = 14), # changes legend title
      legend.text = element_text(size = 10)) + # changes legend text

  labs(x = 'Chemical Category', y = 'Biomass', 
       fill = "Scaled Chemical\nConcentration (pg/mL)") + # changing axis labels 
  scale_fill_gradient(low = rcolorbrewer_colors[1], high = rcolorbrewer_colors[2]) # changing the colors

Module3_2_TYK_figure
```

```{r}
Output = "/Users/alexis/Library/CloudStorage/OneDrive-UniversityofNorthCarolinaatChapelHill/CEMALB_DataAnalysisPM/Projects/P1006. Data Organization & File Management/P1006.2. Modules/P1006.2.4. Improving Data Visualizations/Module3_2"

# exporting figure
ggsave(Module3_2_TYK_figure, 
       filename = 'Module3_2_TYK_figure.png',
       path = Output,
       width = 16, height = 4)
```